

```{r}
y <- c(20, 17, 17, 21, 16, 14, 17, 16, 15, 8, 11, 8)
y <- matrix(y, nrow=length(y))

n <- length(y)
ones <- rep(1,n)
g0 <- c(1, 1, 1)
notg0 <- c(0, 0, 0)
lastg0 <- c(-1,-1,-1)

x1 <- c(g0, notg0, notg0, lastg0)
x2 <- c(notg0, g0, notg0, lastg0)
x3 <- c(notg0, notg0, g0, lastg0)

X <- matrix(c(ones, x1, x2, x3), nrow=n)
```


```{r}

b <- solve(t(X) %*% X) %*% t(X) %*% y
b

yhat <- X %*% b
yhat.ctr <- yhat - mean(yhat)
len.yhat <- t(yhat.ctr) %*% yhat.ctr
dim.yhat <- 3

e <- y - yhat
len.e <- t(e) %*% e
dim.e <- 8

F <- (dim.e * len.yhat)/(dim.yhat * len.e)
F
```

Here's the how to do this in R:

```{r}

groups <- c(1,1,1, 2,2,2, 3,3,3, 4,4,4)
group.factor <- as.factor(groups)

lm.dummy <- lm(y ~ group.factor)
anova(lm.dummy)
summary(lm.dummy)
```

The ANOVA table looks the same as what I presented in the slides, but the model coefficients don't look the same because by default `lm()` uses dummy coding. We use the `contrasts` function to see how R creates contrasts from our grouping variable. 

```{r}
contrasts(group.factor)
```

To see the actual matrix, X, that `lm()` creates to fit the linear model, use `model.matrix()`:

```{r}
model.matrix(~group.factor) # note the use of tilde (see below)
```

We can also use `contrasts()` in combination with `contr.sum` to change this to effect coding. The argument to `contr.sum()` should be the total number of groups:

```{r}
contrasts(group.factor) <- contr.sum(4)
contrasts(group.factor)

anova.2 <- aov(y ~ group.factor)
summary(anova.2)
coefficients(anova.2)
```

```{r}
model.matrix(~group.factor)  # see how it's changed
```

```{r}
lm.effects <- lm(y ~ group.factor)
anova(lm.effects)
summary(lm.effects)
```


```{r}

Y <- matrix(y)
X <- model.matrix(~group.factor)  
   # will be effect coding if you used the contr.sum function above
   # otherwise will be dummy coding
X

b <- solve(t(X) %*% X) %*% t(X) %*% Y
b

yhat <- X %*% b
yhat.ctr <- yhat - mean(yhat)
len.yhat <- t(yhat.ctr) %*% yhat.ctr
dim.yhat <- 3

e <- Y - yhat
len.e <- t(e) %*% e
dim.e <- 8

F <- (dim.e * len.yhat)/(dim.yhat * len.e)
F
```

