

\section{Writing Functions in Python}

The general form of a Python function is as follows:
%
\begin{python}
def funcname(arg1,arg2):
    # one or more expressions
    return someresult # arbitrary python object (could even be another function)
\end{python}
%
An important thing to remember when writing functions is that Python is
white space sensitive. In Python code indentation indicates scoping
rather than braces. Therefore you need to maintain consistent
indentation. This may surprise those of you who have extensive
programming experience in another language. However, white space
sensitivity contributes significantly to the readability of Python code.
Use a Python aware programmer's editor and it will become second nature
to you after a short while. I recommend you set your editor to
substitute spaces for tabs (4 spaces per tab), as this is the default
convention within the python community.

Here's an example of defining and using a function in the Python
interpreter:
%
\begin{python}
>>> def mypyfunc(x,y):
...     return x**2 + y**2 + 3*x*y
... 
>>> mypyfunc(10,12)
604
>>> a = numpy.arange(1,5,0.5)
>>> b = numpy.arange(2,6,0.5)
>>> mypyfunc(a,b)
array([  11.  ,   19.75,   31.  ,   44.75,   61.  ,   79.75,  101.  ,
        124.75])
>>> a = range(1,5)
>>> b = range(1,5)
>>> mypyfunc(a,b)

Traceback (most recent call last):
  File "<pyshell#52>", line 1, in -toplevel-
    mypyfunc(a,b)
  File "<pyshell#45>", line 2, in mypyfunc
    return x**2 + y**2 + 3*x*y
TypeError: unsupported operand type(s) for ** or pow(): 'list' and 'int'
>>> 
\end{python}
%
Note that this function works for numeric types (\lstinline!ints! and
\lstinline!floats!) as well as \lstinline!numpy.arrays! but not for
simple Python lists. If you wanted to make this function work for lists
as well you could define the function as follows:
%
\begin{python}
>>> def mypyfunc(x,y):
...     x = numpy.array(x)
...     y = numpy.array(y)
...     return x**2 + y**2 + 3*x*y
... 
>>> a
[1, 2, 3, 4]
>>> b
[1, 2, 3, 4]
>>> mypyfunc(a,b)
array([ 5, 20, 45, 80])
\end{python}

\subsection{Putting Python functions in Modules}

As with R, you can define your own Python modules that contain user
defined functions. Using a programmer's text editor, write your
function(s) and save it to a file with a \lstinline!.py! extension in a
directory in your |PYTHONPATH| (see below).

\begin{python}
# functions defined in vecgeom.py
import numpy

def veclength(x):
    """Calculate length of a vector x.""" 
    x = numpy.array(x)
    return numpy.sqrt(numpy.dot(x,x))


def unitvector(x):
    """Return a unit vector in the same direction as x."""
    x = numpy.array(x)
    return x/veclength(x)
\end{python}
%
To access your function use an \lstinline!import! statement:
%
\begin{python}
>>> import vecgeom
>>> x = [-3,-3,-1,-1,0,0,1,2,2,3]
>>> help(vecgeom.veclength)
Help on function veclength in module vecgeom:

veclength(x)
    Calculate length of a vector x.

>>> vecgeom.veclength(x)
6.164414002968976
# import all fxns from the vecgeom module
>>> from vecgeom import * 
>>> print vecgeom.unitvector(x)
[-0.48666426 -0.48666426 -0.16222142 -0.16222142  0.          0.   
     0.16222142  0.32444284  0.32444284  0.48666426]
\end{python}



\section{Setting the PYTHONPATH}

Like the operating system, Python searches a set of default directories whenever you ask it to load a specific module.  Python knows where to find all of it's base modules, and a well written package will install it's files into one of the standard locations.

To see the directories that your Python installation searches by default try the following commands in the Python interpreter (your output will be different):
%
\begin{python}
>>> import sys
>>> import sys
>>> sys.path
['', '/Library/Frameworks/Python.framework/Versions/7.1/bin', '/Users/pmagwene', 
'/Users/pmagwene/synchronized/pyth', '/Users/pmagwene/pytest', 
... output truncated ...
\end{python}

For your own code it's useful to setup a separate directory. Create a directory called |pycode| in your home directory. In order for Python to "see" the code in this directory you must add it to your |PYTHONPATH|.

To temporarily add a new directory to |sys.path|:
\begin{python}
>>> sys.path.append('/Users/pmagwene/pycode') # substitute the path to the directory you used
>>> print sys.path
['', '/Library/Frameworks/Python.framework/Versions/7.1/bin', '/Users/pmagwene', 
'/Users/pmagwene/synchronized/pyth', '/Users/pmagwene/pytest', 
... output truncated...
'/Users/pmagwene/pycode']
\end{python}

This change applies only to your current interpreter and lasts until you close the interactive prompt. To make persistent changes to the Python seach path you need to create an environment variable called |PYTHONPATH| and add the desired directories.  You do this the same way you set your system |PATH|, but modify your shell initialization file (Unix or OS X) or using the System Properties tool in the control panel to create a new environment variable (Windows). For example, on OS X add the following line to your |.bash_profile| (found in your home directory, create it if doesn't already exist):
%
\begin{bash}
export PYTHONPATH=$PYTHONPATH:$HOME/pycode   
\end{bash}
%
If you are keeping your Bio723 code somewhere other than |~/pycode| than change the location as needed. To have this change take effect, start a new Terminal window or re-start \ipy.

For more info on setting |PATH| variables see: \url{https://github.com/pmagwene/Bio313/wiki/setting-paths}.